Non-word repetition in Yélî Dnye

### 2021-09-13 manuscript.Rmd
in progress

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong effect on NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

1. Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
2. Sounds (and sound sequences) that are harder to prononunce tend to be mispronounced and thus lost diachronically
3. Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically

@@

#### Individual variation effects on NWR

The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that consistent individual variation effects may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.

Our focus on age stems from previous work, where performance increases with child age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Although past research has not investigated potential effects of birth order on NWR, there is a sizable literature on these effects in other language tasks [e.g., @havron2019effect], and therefore we report on these too. Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation [@lehmann2018early]. Regarding child sex, no such effect has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed. Finally, prior research varies on whether significant differences as a function of  maternal education are reported [e.g., no difference found in @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio; whereas differences were reported in @tuller2018identifying; @dos2020epreuve]. In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019]. The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children [see discussion in @cristia2020infant].

@@

Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic effects (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural effects (i.e., potential differences in NWR as a function of other differences across human populations).^[Please note that the linguistic and cultural effects discussed here are different from the effects discussed in the extensive literature on NWR by bilingual participants. In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment. To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge. For instance, @tuller2018identifying employed a set of non-words judged to be language independent and two others that were more aligned with either French or German. The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language ("language-dependent") than when they are "language-independent". To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience. Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR [@farabolini2021association]. In any case, this line of research focuses on links between exposure to a given language and NWR performance. In contrast, when we discuss linguistic or cultural effects, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation. Specifically, we investigated potential effects of age, child sex, maternal education, and birth order. There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.

@@

Some general ideas regarding potential maternal education effects on our data may be drawn from the observations above. To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken maternal education effects. 
<!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years). ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment -->
In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.

As for birth order, much of the work on birth order effects on cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology]. It is possible that birth order effects are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation. This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.

@@

We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them. However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known effects, given the previously discussed variability across studies. Therefore, all analyses in the present study are descriptive and should be considered exploratory.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-46 items: @jabere2018xperiment; @kalnak2014nonword;@piazzalunga2019articulatory], with the entire test inventory of 40 final items distributed across children. We  used a relatively large number of items to explore effects of length and phonological complexity. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children. Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this effect.

@@

Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = `r mycoeff["avg_fr_s:age.rounded_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s:age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for older children. Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.

@@

### Individual variation effects on NWR

@@

We do not have the necessary data to assess whether the effect is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often. This effect was large, with a magnitude more than twice the size of the effect of participant age. Moreover, this significant effect remained even  after accounting for the frequencies of these segments in a conversational corpus. An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.

We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development. We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this effect is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR. We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop [as in e.g., @edwards2004interaction].

@@

In contrast, previous work shows little evidence for effects of maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]  on NWR scores. We did not expect large effects of maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. We find effects of about that size for participant sex, which is aligned with previous work [@chiat2007preschool].

Finally, we investigated whether birth order might affect NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings. Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers. Therefore we expected that any effects of birth order on NWR would be attenuated in this context. In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score. However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`. In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the effects we found are larger. We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.

>>>>

abstract: >
   In non-word repetition (NWR) studies, participants are presented auditorily with an item that is phonologically legal but lexically meaningless in their language, and asked to repeat this item as closely as possible. NWR scores are thought to reflect some aspects of phonological development, saliently a perception-production loop supporting flexible production patterns. In this study, we report on NWR results among children (N= 40, aged 3-10 years) learning Yélî Dnye, an isolate spoken on Rossel Island in Papua New Guinea. Results make three contributions that are specific, and a fourth that is general. First, we found that non-word items containing typologically frequent sounds are repeated without changes more often that non-words containing typologically rare sounds, above and beyond any within-language frequency effects. Second, we documented rather weak effects of item length. Third, we found that age has a strong correlation with NWR scores, whereas there are weak correlations with child sex, maternal education, and birth order. Fourth, we weave our results with those of others to serve the general goal of reflecting on how NWR scores can be compared across participants, studies, languages, and populations,  and the extent to which they shed light on the factors universally structuring variation  in phonological development at a global and individual level.

@@

1. Sounds (and sound sequences) that are harder to perceive tend to be misperceived and thus lost diachronically
2. Sounds (and sound sequences) that are harder to pronounce tend to be mispronounced and thus lost diachronically
3. Sound sequences that are harder to hold in memory tend to be mispronounced and thus lost diachronically

@@

#### Individual variation correlations with NWR

The third research area we contribute data to relates to the possibility that children differ from each other in NWR scores in systematic ways. Although the ideal systematic review is missing, a recent paper comes close with a rather extensive review of the literature looking at correlations between NWR scores and a variety of child-level variables, including familial socio-economic status, child vocabulary, and, among multilingual children, levels of exposure to the language on which the non-words are based [@farabolini2021nonword]. In a nutshell, most evidence is mixed, suggesting that correlations with individual variation may be small, and more data is needed to estimate their true size. For this reason, we descriptively report association strength between NWR scores and child age, sex, birth order, and maternal education.

Our focus on age stems from previous work, where performance increases with child age [@farmani2018normalization; @kalnak2014nonword; @vance2005speech]. Although past research has not investigated potential correlations with birth order on NWR, there is a sizable literature on these correlations in other language tasks [e.g., @havron2019effect], and therefore we report on these too. Common explanations for advantages for first- over later-born children include differential allocation of familial resources, particularly parental behaviors of cognitive stimulation [@lehmann2018early]. Regarding child sex, no significant correlation has been found in previous NWR research [@chiat2007preschool], and in other language tasks evidence is mixed. Finally, prior research varies on whether significant differences as a function of  maternal education are reported [e.g., no difference found in @farmani2018normalization; @kalnak2014nonword; @meir2017independent; @balladares2016socio; whereas differences were reported in @tuller2018identifying; @dos2020epreuve]. In other lines of work, maternal education correlates with child language outcomes, including vocabulary reports [@frank2017wordbank] and word comprehension studies [@scaff2019]. The causal pathways explaining this correlation are complex, but one explanation that is often discussed involves more educated mothers talking more to their children [see discussion in @cristia2020infant].

@@

Aside from @cristia2020infant's hypotheses just mentioned, we have found little discussion of linguistic differences (i.e., potential differences in NWR as a function of which specific language children are learning, and/or its typology) or cultural differences (i.e., potential differences in NWR as a function of other differences across human populations).^[Please note that the linguistic and cultural differences discussed here are different from the differences discussed in the extensive literature on NWR by bilingual participants. In that literature, authors are concerned with individual variation in exposure to one (as opposed to other) languages among multilingual children, as variation in relative language experiences could mask potential effects of language impairment. To try to measure language abilities above and beyond relative levels of experience with a given language, authors have tried to build non-words that tap language-dependent or language-independent knowledge. For instance, @tuller2018identifying employed a set of non-words judged to be language independent and two others that were more aligned with either French or German. The intuition is that NWR will correlate with the relative levels of exposure to that language bilingual children more strongly when items are aligned with a specific language ("language-dependent") than when they are "language-independent". To make this more precise, among bilingual children, those that have more experience with English than Spanish should perform better on English non-words than their peers with less English experience. Preliminary results of an ongoing meta-analysis suggest significant associations between exposure to a given language and performance in both language-dependent and language-independent NWR [@farabolini2021association]. In any case, this line of research focuses on links between exposure to a given language and NWR performance. In contrast, when we discuss linguistic or cultural differences, we ask the question of whether children vary in their performance as a function of which language they are learning and/or their overall levels of language experience (not relative levels in a multilingual setting).]

@@

Some aspects of the community are relevant for contextualizing our study design and results, particularly regarding sources of individual variation. Specifically, we investigated potential correlations with age, child sex, maternal education, and birth order. There is nothing particular to note regarding age and child sex, but we have some comments that pertain to the other two factors.

@@

Some general ideas regarding potential correlations between our NWR measures and maternal education may be drawn from the observations above. To begin with, many of our participants above 6 years of age may not be living with their birth mother but with other relatives, which may weaken associations with maternal education. 
<!-- Additionally, the importance given to formal education appears relatively stable over the period that Rossel Island has been visited by language researchers (Steven Levinson and Penelope Brown, about 20 years). ac2mc removed, R was confused and I'm also not sure what is the relevance of this comment -->
In addition, it seems to us that the length of formal education a given individual may have is not necessarily a good index of their socio-economic status or other individual properties, unlike what happens in industrialized sites, and variation may simply be due to random factors like living close to a school or having relatives there.

As for birth order, much of the work on correlations between birth order and cognitive development (including language) has been carried out in the last 70 years and in agrarian or industrialized settings [@gratz2018competition;@barclay2015within], where nuclear families are more likely to be the prevalent rearing environment [@lancy2015anthropology]. It is possible that birth order differences are stronger in such a setting, because much of the stimulation can only come from the parents, and when there are multiple children, the inter-birth interval is small enough that older siblings may not be of an age that allows them to contribute to their younger siblings' stimulation. This contrasts with the picture just drawn in the Yélî community, where children will typically benefit from a rich and extensive socially stimulating setting, surrounded by siblings, and cousins of several orders, regardless of their birth order in their nuclear family.

@@

We had considered boosting the interpretational value of this evidence by announcing our analysis plans prior to conducting them. However, we realized that even pre-registering an analysis would be equivocal because we would not have enough power to look at all relationships of interest, in many cases possibly not enough to detect any of the known associations, given the previously discussed variability across studies. Therefore, all analyses in the present study are descriptive and should be considered exploratory.

@@

Many NWR studies are based on a fixed list of 12-16 items that vary in length between 1 and 4 syllables, often additionally varying syllable complexity and/or cluster presence and complexity, and always meeting the condition that they do not mean anything in the target language [e.g., @balladares2016socio; @wilsenach2013phonological]. We kept the same variation in item length and requirement for not being meaningful in the language, but we did not vary syllable complexity or clusters because these are vanishingly rare in Yélî Dnye. We also increased the number of items an individual child would be tested on, such that a child would get up to 23 items to repeat [other work has also used up to 24-46 items: @jabere2018xperiment; @kalnak2014nonword;@piazzalunga2019articulatory], with the entire test inventory of 40 final items distributed across children. We  used a relatively large number of items to explore correlations with length and phonological complexity. However, aware that this large item inventory might render the task longer and more tiresome, we split items across children. Naturally, designing the task in this way may make the study of individual variation within the population more difficult because different children are exposed to different items.

@@

We also looked into the frequency with which mispronunciations resulted in real words. In fact, two thirds of incorrect repetitions were recognizable as real words or phrases in Yélî Dnye or English: `r round(sum(!is.na(trnsc.inc$meaning) & trnsc.inc$meaning != "0")/sum(!is.na(trnsc.inc$mispronunciation))*100)`%. This type of analysis is seldom reported. We could only find one comparison point: @castro1998illiterate found that illiterate European Portuguese adults' NWR mispronunciations resulted in real words in `r round(53/475*100,2)`% of cases, whereas literate participants did so in only `r round(2/117*100,2)`% of cases. The percentage we observe here is much higher than reported in the study by Castro and colleagues, but we do not know whether age, language, test structure, or some other factor explains this difference, such as the particularities of the Yélî Dnye phonological inventory, which lead any error to result in many true-word phonetic neighbors. Follow-up work exploring this type of error in children from other populations in addition to further work on Yélî children may clarify this association.

@@

Additionally, the effect for the interaction between the two fixed effects was small but significant (ß = `r mycoeff["avg_fr_s:age.rounded_s","Estimate"]`, SE ß = `r mycoeff["avg_fr_s:age.rounded_s","Std. Error"]`, p `r pval.display(mycoeff["avg_fr_s:age.rounded_s","Pr(>|z|)"])`): The effect of frequency was larger for younger children. Inspection of Figure \@ref(fig:fig-xling-freq-age3) suggests that the age effects are more marked for items containing cross-linguistically common phones, such that children's average performance increases more rapidly with age for those than for items containing cross-linguistically common phones.

@@

### Individual variation and NWR

@@

We do not have the necessary data to assess whether the correlation is indeed weaker for Yélî Dnye learners than learners of other languages, but we did find a robust correlation of average segmental cross-linguistic frequency and NWR performance: Even accounting for age and random effects of item and participant, we saw that target words with typologically more common segments were repeated correctly more often. This effect was large, with a magnitude more than twice the size of the effect of participant age. Additionally, we observed an interaction between age and this factor, which emerged because cross-linguistic frequency explained more variance at younger ages (i.e., the difference in performance for more versus less typologically frequent sounds was greater for younger than older children). Importantly, the correlation between performance and typological frequencyis  remained significant  after accounting for the frequencies of these segments in a conversational corpus. An analysis of the substitutions made by children also aligned with this interpretation, with typologically more common sounds being substituted for typologically less common ones.

We thus at present conclude that typological frequency of sounds is, to a certain extent, mirrored in children's NWR, in ways that may not be due merely to how often those sounds are used in the ambient language, and which are not erased by language-specific pressure to make finer-grained differences early in development. We do not aim to reopen a debate on the extent to which cross-linguistic frequency of occurrence can be viewed necessarily as reflecting ease of perception or production [most often discussed in the case of phonotactic constraints on sequences, e.g., @maddieson2009phonology], but we do point out that this association is interestingly different from effects found in artificial language learning tasks [see @moreton2012structure for a review] which are in some ways quite similar to NWR. We believe that it may be insightful to extend the purview of NWR from a narrow focus on working memory and structural factors to broader uses, including for describing the phonological representations in the perception-production loop [as in e.g., @edwards2004interaction].

@@

In contrast, previous work varies with respect to correlations of NWR scores with maternal education [e.g., @farmani2018normalization; @kalnak2014nonword; @meir2017independent]. We did not expect large correlations with maternal education in our sample for two reasons: First, education on Rossel Island is generally highly valued and so widespread that little variation is seen there; second, formal education is not at all essential to ensuring one's success in society and may not be a reliable index of local socioeconomic variation locally. In fact, maternal education correlated with NWR score at about r~.1, which is small. We find correlations of about that size for participant sex, which is aligned with previous work [@chiat2007preschool].

Finally, we investigated whether birth order might correlate with NWR scores, as it does other language tasks, resulting in first-born children showing higher scores on standardized language tests than later-born children [@havron2019effect] and adults [in a battery including verbal abilities, e.g., @barclay2015within], presumably because later-born children receive a smaller share of parental input and attention than their older siblings. Given shared caregiving practices and the hamlet organization typical of Rossel communities, children have many sources of adult and older child input that they encounter on a daily basis and first-born children quickly integrate with a much larger pool of both older and younger children with whom they partly share caregivers. Therefore we expected that any correlations with birth order on NWR would be attenuated in this context. In line with this prediction, our descriptive analysis showed a non-significant correlation between birth order and NWR score. However, the effect size was larger than that found for the other two factors and it is far from negligible, at r~.2 or Cohen's d~`r r2d(.2)`. In fact, two large studies with therefore precise estimates found effects of about d~.2 [@havron2019effect;@barclay2015within], which would suggest the correlations we found are larger. We therefore believe it may be worth revisiting this question with larger samples in similar child-rearing environments, to further establish whether distributed child care indeed results in more even language outcomes for first- and later-born children.
